---
title: "Extending GAMs with interactions"
author: "Eric Pedersen"
date: "November 4, 2021"
output:
  xaringan::moon_reader:
    css: ['default', 'https://stackpath.bootstrapcdn.com/bootstrap/4.3.1/css/bootstrap.min.css', 'slides.css']
    lib_dir: libs
    nature:
      titleSlideClass: ['inverse','middle','left',my-title-slide]
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      beforeInit: "macros.js"
      ratio: '16:9'
---

```{r setup, include=FALSE, cache=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(cache = TRUE, dev = 'svg', echo = TRUE, message = FALSE, warning = FALSE,
                      fig.height=6, fig.width = 1.777777*6)
library('here')
library('mgcv')
library('gratia')
library('gamair')
library('ggplot2')
library('purrr')
library('mvnfast')
library("tibble")
library('gganimate')
library('tidyr')
library("knitr")
library("viridis")
library('readr')
library('dplyr')
library('gganimate')
library('xaringan')
library('patchwork')
library('faraway')
library('sf')
library(extrafont)
## plot defaults
theme_set(theme_minimal(base_size = 16, base_family = 'Fira Sans'))
## constants
anim_width <- 1000
anim_height <- anim_width / 1.77777777
anim_dev <- 'png'
anim_res <- 200
```

# Reminder from Tuesday:

A GLM is: 
  - a distribution for our data + 
  - a link function + 
  - a linear model of the transformed mean as a function of covariates

--

A GAM is a GLM except:
  - covariate functions can be nonlinear, built from basis functions of covariates
  - Estimated with penalized likelihood, penalizing overly wiggly functions
  
  
---


# So far:

- One-dimensional smooth models
- Models with multiple separate predictors
- Normally distributed response

These are good models, but there's lots more to see!


---

# New things

- distribution of the data
  - `family=` argument
- visually checking model assumptions when building models
- adding dimensions to your smooths
  - space and time


---

# Distributions

- `family=` argument in `gam()`
- see `?family.mgcv` for a list
- what you'd expect, as for `lm()` and `glm()`
- most useful cases:
  - `binomial` ( yes/no, $y \in \{0, 1\}$ )
  - `poisson` ( counts, $y \in \{0, 1, 2, \ldots\}$ )
  - `Gamma` ( positive, $y>0$ )


---

# Special count distributions

- Poisson is often not adequate for "real" counts
- Assuming $\text{Var}(y) = \mathbb{E}(y)$ is usually incorrect
 

```{r species-gala-plots, echo=FALSE, eval = TRUE,fig.height=4}


species_model <- gam(Species ~ log(Area), family=poisson, data=gala)

species_model_plot <- gala%>%
  ggplot(aes(x = Area, y= Species))+
  geom_point()+
  geom_line(aes(y=  fitted(species_model)))+
  scale_x_log10()+
  labs(title = "Galapagos Island SAR\nSpecies ~ log(Area)\nfamily=Poisson")


species_model_plot
```

---

# Special count distributions

- Poisson is often not adequate for "real" counts
- Assuming $\text{Var}(y) = \mathbb{E}(y)$ is usually incorrect
 
```{r species-gala-plots2, echo=FALSE, eval = TRUE, fig.height=4}
species_resid_plot <- gala %>%
  mutate(`Predicted mean` = fitted(species_model),
         `Predicted Var` = `Predicted mean`, 
         `Observed residuals` = Species-`Predicted mean`,
         `Simulated residuals` = rpois(n = n(), lambda = `Predicted mean`)- `Predicted mean`)%>%
  select(`Predicted mean`,`Predicted Var` ,`Observed residuals`, `Simulated residuals`)%>%
  gather(key = "outcome", value= "residual", -`Predicted mean`, - `Predicted Var`)%>%
  mutate(outcome =factor(outcome, levels = c("Simulated residuals","Observed residuals")))%>%
  ggplot(aes(x= `Predicted mean`, y= residual))+
  facet_grid(.~outcome)+
  geom_point()+
  geom_line(aes(y =sqrt(`Predicted Var`)))+
  geom_line(aes(y =-sqrt(`Predicted Var`)))
  
species_resid_plot

```
---

# Special count distributions


Other options?
 
- `quasipoisson` (count-ish, $y \geq 0$)
  - $\text{Var}(y) = \psi\times\mathbb{E}(y)$ 
  - awkward to check, no likelihood
  
--

- `nb`/`negbin` ( $y \geq 0$)
  - $\text{Var}(y) = \mathbb{E}(y) + \kappa\times\mathbb{E}(y)^2$ 
  - models overdispersed counts
  
--

- `tw`/`Tweedie` ( $y \geq 0$)
  - great for CPUE-type data
  - works on non-count positive data as well
  - $\text{Var}\left(\text{y}\right) = \phi\mathbb{E}(\text{y})^q$

---

# Tweedie distribution

.pull-left[
```{r tweedie, echo=FALSE, fig.height=5, fig.width=5}
par(ps=17)
library(RColorBrewer)
library(mgcv)

# taken from ?mgcv::ldTweedie
y <- seq(1e-10,10,length=1000)
p <- seq(1.2, 1.6, by=0.2)
phi <- 1
fy <- exp(ldTweedie(y,mu=2,p=p[1],phi=phi)[,1])

rr <- brewer.pal(8,"Dark2")
  
plot(y, fy, type="n", ylim=c(0,1), main="Tweedie density as q changes",
     xlab="Count", ylab="Density")
for (i in 1:length(p)) {
  fy <- exp(ldTweedie(y, mu=2, p=p[i], phi=phi)[, 1])
  lines(y, fy, col=rr[i], lwd=2.5)
}

legend(x="topright", legend=p, fill=rr, inset=0.05)

```
(NB there is a point mass at zero not plotted)
]
.pull-right[
-  $\text{Var}\left(\text{y}\right) = \phi\mathbb{E}(\text{y})^q$
- Poisson is $q=1$ 
- We estimate $q$ and $\phi$ for `tw`
- We set $q$ and estimate $\phi$ for `Tweedie`
]

---

# Negative binomial distribution

.pull-left[
```{r negbin, echo=FALSE, fig.height=5, fig.width=5}
y<-seq(1,12,by=1)
disps <- seq(0.01, 1, len=5)

fymat <- matrix(NA, length(y), length(disps))

i <- 1
for(disp in disps){
  fymat[,i] <- dnbinom(y, size=disp, mu=5)
  i <- i+1
}

plot(range(y), range(fymat), type="n", ylab="Density", xlab="Count", cex.lab=1.5,
     main="")

rr <- brewer.pal(8,"Dark2")

for(i in 1:ncol(fymat)){
  lines(y, fymat[,i], type="l", col=rr[i], lwd=2)
}

legend(x="topright", legend=disps, fill=rr, inset=0.05)

```
]
.pull-right[
- $\text{Var}\left(\text{count}\right) =$ $\mathbb{E}(\text{count}) + \kappa \mathbb{E}(\text{count})^2$
- Poisson is $\kappa=0$
- Estimate $\kappa$ for `nb`
- Set $\kappa$ for `negbin` 
]

---

# 游游멇릟 Example 游냐游멇릟

.pull-left[
- Let's look at species richness as a function of depth
- Number of species in each trawl (counts!)

```{r richness}
trawls <- read.csv(here("data/trawl_nl.csv"))
```
]

.pull-right[
```{r richness-violin, fig.height=5, fig.width=5, echo=FALSE}
ggplot(trawls, aes(x =depth, y=richness)) +
  geom_point()+
  scale_y_log10()+
  labs(x="Depth (m)", y="Number of Species")
```

]

---

# 游游멇릟 Example 游냐游멇릟

.pull-left[
- Does richness vary with depth?
- Let's try assuming the response is Poisson!

```{r pois-rich-fit}
rich_depthl10 <- gam(richness~s(log10(depth),k = 30), 
                     family=poisson,
              method="REML", 
              data=trawls)
```
]

.pull-right[
```{r plot-year-rich, fig.width=5}
plot(rich_depthl10, shade=TRUE)
```
]




---

# `summary` output

```{r rich-sum, echo=FALSE}
summary(rich_depthl10)
```

---

# Other `family` stuff

- can specify a `link=` argument
  - (usually don't have to)
  - `?family` has options for each distribution
- for a fitted model `model$family`
  - details of what was used
  - `model$family$linkfun()` gives link function
  - `model$family$linkinv()` gives inverse link

---

# How can we check whether our assumed distribution is adequate?


```{r pois-rich-check1, fig.height=4}
par(mfrow=c(2,2))
gam.check(rich_depthl10)
```

---


# How can we check whether our assumed distribution is adequate?


```{r pois-rich-check2,fig.show="hide"}
par(mfrow=c(2,2))
gam.check(rich_depthl10)
```

---

# How can we check whether our assumed distribution is adequate?

```{r pois-rich-check3, fig.height=4}

appraise(rich_depthl10)

```

---

--- 

# Let's pause the talk here to try this ourselves

---

class: inverse middle center big-subsection

## Beyond one dimensional smooths:<br/> space, time and more

---

# 游붏游붏 Example 游붏游붏

.pull-left[
```{r data-load}
trawls <- read.csv(here("data/trawl_nl.csv"))
trawls_2010 <- filter(trawls, year==2010)
range(trawls_2010$shrimp)
head(trawls_2010$shrimp, 3)
```

Not normal, not integer!

]

.pull-right[

```{r, fig.height=5, fig.width=5, echo=FALSE}
hist(trawls_2010$shrimp, xlab="Biomass", main="")
```

]

---

# 游붏游딬

.pull-left[
- spatial variation!
- how do we model this??
- `s(long) + s(lat)` misses the interaction
]

.pull-right[
```{r biom-space-plot, fig.height=11, echo=FALSE}
coast <- read_sf(here("data/nl_coast.shp"))

ggplot(trawls_2010) +
  geom_point(aes(x=long, y=lat, size=shrimp)) +
  geom_sf(data=coast) +
  coord_sf()
```
]

---

# Adding dimensions

- Thin plate regression splines (default basis)
- `s(x, y)`
- Assumes `x` and `y` are measured in the same units
  - `x`, `y` projected coordinates 游녨
  - `x` temperature, `y` depth 游녩

---

# 游붏游딬

Starting with a 2D smooth, assuming data is Gaussian


.pull-left[
```{r shrimpmod}
spatial_shrimp <- gam(shrimp ~ s(x, y, k =100),
                data=trawls_2010,
                family = gaussian,
                method="REML")
```
]

.pull-right[
```{r plot-effect, fig.height=8}
plot(spatial_shrimp, asp=1)
```
]

---

# That plot is hard to understand!

.pull-left[
```{r plot-effect-bad, fig.height=10}
plot(spatial_shrimp, asp=1)

```
]

.pull-right[
```{r plot-effect-good, fig.height=10}
plot(spatial_shrimp, scheme=2, asp=1)

```
]

---

# Using gratia:
.left-column[

```{r plot-effect-gratia, fig.height=5, fig.show='hide'}
draw(spatial_shrimp)

```

]

.right-column[
```{r plot-effect-gratia2, fig.height=8, echo=FALSE}
draw(spatial_shrimp)

```
]

---

# `summary` output

```{r summary-spat, echo=FALSE}
summary(spatial_shrimp)

```

---

class: inverse middle center big-subsection

## Now give it a try!


---

class: inverse middle center big-subsection

## Other multi-dimensional smooths

---

# `s(x,y)` doesn't always work

- Only works for `bs="tp"` or `bs="ts"`
- Covariates are isotropic
- What if we wanted to use lat/long?
- Or, more generally: interactions between covariates?

---

# Enter `te()`

.pull-left[
- We can built interactions using `te()`
- Construct 2D basis from 2 1D bases
- Biomass as a function of temperature and depth?
  - `te(temp_bottom, log10(depth))`
- 游눬 "marginal 1Ds, join them up"
]

.pull-right[
```{r tensor, echo=FALSE, results='hide', fig.width=8, fig.height=6, messages=FALSE, warning=FALSE}
b_shrimp_ll_ti <- gam(shrimp ~ ti(log10(depth), temp_bottom ) + ti(depth) + ti(temp_bottom)  ,
                data=trawls_2010,
                family=tw,
                method="REML")

layout(matrix(c(1,2,3,3), 2, 2), widths=c(1.5,1.5,1), height=c(1,1,2))
opar <- par(mar=c(4, 3, 1, 2) + 0.1)
plot(b_shrimp_ll_ti, select=3)
plot(b_shrimp_ll_ti, select=2)
par(mar=c(0, 0, 0, 0) + 0.1)
vis.gam(b_shrimp_ll_ti, view=c("temp_bottom","depth"), theta=-60, phi=30, color="bw")
par(opar)
```
]

---

# Using `te()`

Just like `s()`:


```{r te-ex}
shrimp_te <- gam(shrimp ~ te(log10(depth), temp_bottom),
                data=trawls_2010,
                family=tw,
                method="REML")
```

---

# `summary`

```{r te-ex-summary, echo=FALSE}
summary(shrimp_te)
```

---

# Things to fiddle with

- Setting `k` in two ways:
  - `k=5`: 5 for all covariates (total $5*5=25$)
  - `k=c(3,5)`: per basis, in order (total $3*5=15$)
- Setting `bs` in two  ways:
  - `bs="tp"`: tprs for all bases
  - `bs=c("tp", "tp")`: tprs per basis

---

# Pulling `te()` apart: `ti()`

- Can we look at the components of the `te()`
- `te(x, y) = ti(x, y) + ti(x) + ti(y)`

```{r tiex}
shrimp_ti <- gam(shrimp ~ ti(temp_bottom, depth) +
                          s(temp_bottom) + s(depth),
                data=trawls_2010,
                family=tw,
                method="REML")
```

---

# `summary`

```{r ti-ex-summary, echo=FALSE}
summary(shrimp_ti)
```

---

class: inverse middle center big-subsection

## Now give it a try!


---
class: inverse middle center big-subsection

## Building spatio-temporal models

---

# 游붏游딬游딉

```{r biom-space-time-plot, fig.height=8, fig.width=15, echo=FALSE}
ggplot(trawls) +
  geom_point(aes(x=long, y=lat, size=shrimp), alpha=.5) +
  geom_sf(data=coast) +
  facet_wrap(~year, ncol=5)
```


---

# Space x time

- We had a 2d spatial model, add time?
  - `te(x,y,year)` ?
- Want that 2d smooth rather than `te(lat, long)`?
- `d=` groups covariates
  - `te(x, y, year, d=c(2,1))` gives `x,y` smooth and `year` smooth tensor
- Assuming default `k=` and `bs=` for bases

---

# Fiddling

- Often fewer temporal replicates
  - Fewer years than unique locations
  - `k=` smaller for temporal covariate?
- Use cubic spline basis for time?
  - simpler basis, even knot placement
- When using `ti()` everything needs to match up!

---

# 游붏游딬游딉

Putting that together:

```{r ti-xyt}

shrimp_xyt <- gam(shrimp ~ ti(y,x, year, d=c(2,1), 
                              bs=c("tp", "cr"), k=c(20, 5)) +
                           s(x, y, bs="tp", k=20) +
                           s(year,  bs="cr", k=5),
                data=trawls,
                family=tw,
                method="REML")
```

---

# `summary`

```{r xyt-summary, echo=FALSE}
summary(shrimp_xyt)

```

---

# `ti(x, y)` and `ti(year)`

```{r ti-xyt-plot-xy-t, echo=FALSE, results="hide", fig.width=7}
par(mfrow=c(1, 2))
plot(shrimp_xyt, select=2, scheme=2, asp=1)
plot(shrimp_xyt, select=3, scheme=2, scale=0, seWithMean=TRUE)


```
---

# `ti(x, y, year)`

```{r ti-xyt-plot-xyt, echo=FALSE, results="hide", fig.width=7}
plot(shrimp_xyt, select=1, scheme=2, asp=1)

```



---
class: inverse middle center big-subsection

## Recap

---

# What did we learn?

- set response distribution
  - `family=` argument
  - see `?family`
- spatial smoothing
  - `s(x,y)` for projected coordinates
  - `te(lat, long)` for latitude and longitude (better to project?)
- interactions
  - `te(covar1, covar2, ...)`
  - `ti()` to decompose the effects
- space and time
  - `te(x, y, time, d=c(2,1))`
  - again `ti()` to decompose

